Nonlinear Intrinsic Variables and State Reconstruction in Multiscale Simulations

### 2013-12-27 paper/jcp_paper_revision.tex
Added files related to paper

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data \cite{Tenenbaum2000,Roweis2000,Donoho2003,Belkin2003,Coifman2006}.
%
These developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.

While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
 data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by an affine transformation
(see, by analogy, the discussion in Lafon {\em et al.}\cite{lafon2006data}),
the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with, for example, spectral information).
%
Clearly, the ability to systematically construct a {\em unique} and {\em consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities, is invaluable.
%
We will call these coordinates Nonlinear Intrinsic Variables.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements in a test data set from a training data set
that consists of {\em different} observations.
%
To construct this empirical mapping and the associated observers,
accurate interpolation tools must be available in the embedding space; to this
end, we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.

@@

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
%
We impose two critical assumptions. First, the measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t)),
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold,
and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables (with $d \ll n$).
%
Second, the dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d,
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
%
The independence is our second critical assumption.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$ \cite{singer2008non}.
%
Because the empirical method we will describe is independent of the observation function $\mathbf{f}$,
we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
%
The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $\mathbf{f}$ in various observable domains,
or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $\mathbf{f}$
(see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
%
We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model.

The method consists of the following main principles.
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise,
and hence, creates a ``sphere" of samples in the underlying domain $\mathbb{R}^d$.
%%
%%%RONEN
% whether here or in the discussion there should be something about testing this
% practically
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $\mathbf{f}$.
%
In this work, the identification of the associated ellipse of samples according to the time trajectory of our data enables us
to estimate the tangent planes of the observable manifolds {\em via the principal components of the covariance matrices of the samples in these ellipses}.
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $\mathbf{f}$}.
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using this Riemannian metric.

@@

Let $\mathbf{D}$ be a diagonal matrix whose elements are the row sums of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
%
The eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, denoted $\psi_j$, reveal the underlying structure of the data \cite{coifman2005geometric}.
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process.
%
The eigenvectors are ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$, where $\lambda_j$ is the eigenvalue associated with eigenvector $\psi_j$.
%
Because $\mathbf{W}^{\mathrm{norm}} \sim \mathbf{D}^{-1}\mathbf{W} $, and $\mathbf{D}^{-1}\mathbf{W}$ is row-stochastic,
$\lambda_1 = 1$ and $\psi_1$ is the diagonal of $\mathbf{D}^{1/2}$.
%
The next few eigenvectors can be argued to describe the geometry of the underlying manifold \cite{coifman2005geometric}.
%
However, some eigenvectors can be higher harmonics of the same principal direction along the data manifold.
This is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$;
one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

@@

@@

A na\"{\i}ve way to extend $\mathbf{f}$ to a new data point $\Psi(\tilde{i})$ is find the point's nearest neighbors in NIV space and average their function values.
%
A different, point-wise adaptive approach is described by Buchman {\em et al.} \cite{buchman2011high}:
high-dimensional hurricane tracks were estimated from low dimensional embedding coordinates using a weighted average of the points close to $\Psi(\tilde{i})$ in the embedded space.
%
However, this point-wise adaptation requires setting the nearest neighborhood radius parameter for every point.
%
The LP algorithm finds the appropriate nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
%
This radius will be large in smooth regions of the function, and small in regions in which $\mathbf{f}$ contains higher frequency components.
%
The LP approximation of a new, high dimensional point $\mathbf{Y}(\tilde{i})$ is calculated by a weighted average of the function values that belong to the neighboring points.
%
The weights are based on the pairwise distances in the intrinsic, low-dimensional space.
%
In practice, a set of smoothing operators $P_0, P_1, \ldots, P_l$, with
\begin{equation} \label{eq:LP_multi_scale_app}
P_l \propto e^{-d^2(\Psi(i),\Psi(j)) / 2^{-l} \sigma_0},
\end{equation}
are constructed and later extended to create the multiscale approximations as defined in Eq. \ref{eq:LP_multi_scale}.
%
The LP algorithm for the inverse mapping is summarized in Algorithm \ref{algo_LP}.

@@

\item
Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $\mathbf{f}:\Psi(i) \rightarrow Y(i)$.

@@

%\item
%The value $Y(\tilde{i})$ is approximated by $\mathbf{f}_l(\Psi(\tilde{i}))$.

@@

We first consider a chemical reaction network involving  multiple enzyme-substrate interactions \cite{zagaris2012stability}.
%
The reaction steps that comprise the network are\\
\begin{equation}
\begin{array}{rcl}
E + S \overset{e_1}{\underset{e_{-1}}{\leftrightharpoons}} & E:S & \overset{e_2}{\rightarrow} E + S^{*} \\
S + E \overset{b_1}{\underset{b_{-1}}{\leftrightharpoons}} & S:E & \overset{b_2}{\rightarrow} S + E^{*}\\
D + S^{*} \overset{d_1}{\underset{d_{-1}}{\leftrightharpoons}} & D:S^{*} & \overset{d_2}{\rightarrow} D + S\\
F + E^{*} \overset{f_1}{\underset{f_{-1}}{\leftrightharpoons}} & F:E^{*} & \overset{f_2}{\rightarrow} F + E
\end{array}
\end{equation}

The ``$^{*}$'' denotes an activated form of a species, and the ``:'' denotes a complex formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.
%
There are 10 species in this reaction system.
%
However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions
(which we order as $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
%
We consider a parameter regime in which the ODE approximation of this scheme exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
%
Details about the specific parameter values can be found in Appendix \ref{app:rxn}.

@@

We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations
must be non-negative.
%
We evolve each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain a point $\mathbf{Y}(t) \in \mathbb{R}^6$;
according to the time scales calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
but not long enough for the points to converge to a one-dimensional curve or to the final steady state (see Appendix \ref{app:rxn} for more details).
%
In our stochastic simulations, the initial points appear to converge to an approximate two-dimensional manifold
(in expected value, see Figure \ref{fig:rxn_manifolds}).
%
We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points ``on" this apparent two-dimensional manifold.
%
From each manifold point $\mathbf{y} \in \mathcal{Y}$, we run 20 short simulation ``bursts'', each for 0.2 time units.
%
We denote the endpoints from the short simulations as $\mathcal{Y}^{burst}(\mathbf{y})$.

@@

@@

Our second example comes from the molecular dynamics simulation of a small peptide fragment.
%
Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein caricature for simulation studies
\cite{apostolakis1999calculation, bolhuis2000reaction, chekmarev2004long, ma2005automatic, frewen2009exploration, ferguson2011integrating}.
%
We simulate the motion of Ala2 in explicit solvent using the AMBER 10 molecular simulation package \cite{case2008Amber} with an
optimized version \cite{best2009optimized} of the AMBER ff03 force field \cite{duan2003point}.
%
The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}
with periodic boundary conditions, and the particle mesh Ewald method is used for long-range electrostatic interactions \cite{essmann1995smooth}.
%
The simulation is performed at constant volume and temperature (NVT ensemble), with the temperature being maintained at 300~K with a Langevin thermostat \cite{loncharich1992langevin}.
%
Hydrogen bond lengths are fixed using the SHAKE algorithm \cite{ryckaert1977numerical}.
%
The two dihedral angles $\phi$ and $\psi$ are known to parameterize the free energy surface, which contains three important minima (labeled A, B, and C, see Figure \ref{fig:ala_fes}).
%
Our simulations are concentrated around minimum B in the free energy surface, located at $\phi \approx -65^{\circ}$, $\psi \approx 150^{\circ}$.
%
We start many simulations at $10^{\circ}$ away from the minimum, and allow the simulations to each run for 0.1~ps, while recording the configuration of Ala2 every 1~fs (therefore, each trajectory is 100 points long).
%
Configurations are recorded with all atoms except the hydrogens.

@@

We first compare NIV with direct diffusion maps \cite{coifman2005geometric}, an established nonlinear dimensionality reduction technique.
%
We consider 10,000 data points from our simulation $\mathbf{Y}(1), \dots, \mathbf{Y}(10000) \in \mathbb{R}^{30}$; every 100 data points comes from a continuous simulation trajectory.
%
We construct two data sets:
%
$\mathcal{Y}_{even} = \{\mathbf{Y}(t) \text{ restricted to atoms 2, 4, 6, 8, and 10}: t=1, \dots, 10000 \}$,
and $\mathcal{Y}_{odd} = \{\mathbf{Y}(t) \text{ restricted to atoms 5, 7, and 9}: t=1, \dots, 10000 \}$ (see Figure \ref{fig:ala_fes} for the atom indexing).
%
We then compute the NIV and diffusion maps embeddings for $\mathcal{Y}_{even}$ and $\mathcal{Y}_{odd}$;
for NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$.
%at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance
%(as neighboring points in our data set that do not come from the
%same trajectories should not contribute to the estimate of the local covariance).

@@

%\begin{table}
%\begin{tabular}{| r || c | c | c |}
%  \hline
%  & Coordinate 1 & Coordinate 2 & Coordinate 3 \\
%  \hline
%  \hline
%  NLICA & 0.97 & 0.72 & 0.85 \\
%  \hline
%  DM & 0.95 & 0.76 & 0.74 \\
%  \hline
%\end{tabular}
%\caption{Correlation between the embedding coordinates from NLICA and DMAPS using all and just odd coordinates.}
%\end{table}

@@

@@

@@

\begin{thebibliography}{39}%

@@

>>>>

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data \cite{Tenenbaum2000,Roweis2000,Donoho2003,Belkin2003,Coifman2006}.
%
Often, high-dimensional data do not uniformly fill the entire ambient space, but rather lie on a lower-dimensional manifold.
%
Discovering a parameterization of the manifold therefore yields a compact representation of the data, and may lead to efficient analysis and processing schemes.
%
In particular, these developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
Traditionally, state reduction in chemical reaction networks has been done using some {\em a priori} knowledge about separation of time scales,
for example, using the quasi-steady state approximation to reduce a network of chemical reactions\cite{bowen1963singular}.
%
Alternatively, empirical knowledge about the system of interest has been used to develop coarse grained descriptions,
such as grouping atoms of a macromolecule into larger ``beads'' to accelerate simulations\cite{monticelli2008martini, spiga2013electrostatic,izvekov2005systematic, saunders2013coarse}.
%
The use of manifold learning allows for the systematic discovery of data-driven, low-dimensional descriptions that circumvent the need for such {\em a priori} knowledge of the system.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.
%
For example, such reduction coordinates may be the dihedral angles inferred from trajectories of the atoms of a macromolecule.

While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
 data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different bases spanning the hyperplane - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by an affine transformation
(see, by analogy, the discussion in Lafon {\em et al.}\cite{lafon2006data}),
the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with, for example, spectral information).
%

In this paper, we propose a new method to embed data in a low-dimensional space.
%
This method is based on a kernel constructed from a Riemannian metric that is specially designed to exploit the noise/diffusion induced variability of the data in local neighborhoods.
%
By solving the appropriate eigenvector problem, local affinities from the kernel are integrated into a global coordinate system for the data.

We will demonstrate this can produce a {\em unique} and {\em consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities.
%
We will call these coordinates Nonlinear Intrinsic Variables.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements in a test data set from a training data set
that consists of {\em different} observations.
%
To construct this empirical mapping and the associated observers,
accurate interpolation tools must be available in the embedding space; to this
end, we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.

@@

\subsection{Problem Formulation}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
%
We impose two critical assumptions. First, the measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t)),
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold,
and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables (with $d \ll n$).
%
Second, the dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (\mathbf{x}(t)) dt + d w_i(t), \ i=1,\ldots,d,
	\label{eq:dynamics}
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
%
The independence is our second critical assumption.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$ \cite{singer2008non}.
%
Because the empirical method we will describe is independent of the observation function $\mathbf{f}$,
we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
%
The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $\mathbf{f}$ in various observable domains,
or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $\mathbf{f}$
(see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
%
We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model;
this is further discussed and demonstrated in the experimental study.

The method consists of the following main principles.
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise,
and hence, creates a ``sphere" of samples in the underlying domain $\mathbb{R}^d$.
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $\mathbf{f}$.
%
In this work, the identification of the associated ellipse of samples according to the time trajectory of our data enables us
to estimate the tangent planes of the observable manifolds {\em via the principal components of the covariance matrices of the samples in these ellipses}.
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $\mathbf{f}$}.
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using this Riemannian metric.

@@

Let $\mathbf{D}$ be a diagonal matrix whose elements are the row sums of $\mathbf{W}$, and let
\begin{equation}\label{eq:kernel}
\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}
\end{equation}
be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
%
The diagonal entries of $D$ serve as estimates for the density of data around each point.
%
The eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, denoted $\psi_j$, reveal the underlying structure of the data \cite{coifman2005geometric}.
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process.
%
The eigenvectors are ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$, where $\lambda_j$ is the eigenvalue associated with eigenvector $\psi_j$.
%
Because $\mathbf{W}^{\mathrm{norm}} \sim \mathbf{D}^{-1}\mathbf{W} $, and $\mathbf{D}^{-1}\mathbf{W}$ is row-stochastic,
$\lambda_1 = 1$ and $\psi_1$ is the diagonal of $\mathbf{D}^{1/2}$.
%
The next few eigenvectors can be argued to describe the geometry of the underlying manifold \cite{coifman2005geometric}.
%
However, some eigenvectors can be higher harmonics of the same principal direction along the data manifold.
This is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$;
one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

@@

\subsection{Limiting Diffusion Operator} \label{sec:diff_limit}
Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
\begin{equation} \label{eq:Langevin}
	d\mathbf{x}(t) = - \nabla U(\mathbf{x}(t)) dt + d \mathbf{w}(t),
\end{equation}
where the drift term is expressed as a gradient of a potential $U$ and $\dot{\mathbf{w}}(t) \in \mathbb{R}^d$ is a vector consisting of independent white noises.
%
Nadler {\em et al.} \cite{nadler2006diffusion} showed that in the limit $\varepsilon \rightarrow 0$ and $T \rightarrow \infty$ the kernel $\mathbf{W}^{\mathrm{norm}}$ converges to the continuous backward Fokker-Planck operator
\begin{equation}
	\mathcal{L} = \Delta - \nabla U \cdot \nabla.
	\label{eq:limiting_operator}
\end{equation}
%
We remark that, while in diffusion maps \cite{coifman2005geometric}, the limiting operator is the Laplacian on the manifold on which the observations lie,
here the Laplacian operator on the data that is based on the Mahalanobis distance converges to a limiting Laplacian operator on the
manifold of the intrinsic variables $\mathbf{x}$.

The convergence to the limiting operator implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator on the NIV manifold.
%
As such, according to \eqref{eq:limiting_operator}, they depend on the potential.
%
Thus, the constructed embedding is consistent as long as the subsets of observations $\mathbf{Y}(t)$ are
adequate samples of the same (equilibrium) density.
%
For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation sufficiently samples the same underling free energy surface.

In order to merge observations from different regions,
the fundamental question is, to what extent the respective sampling densities affect the coordinates we obtain.
%
In principle, we can decouple diffusion dynamics (and thus, equilibrium densities)
from geometry by normalizing the diffusion kernel with the sample density.
%
This has been explored for diffusion maps independently of NIV \cite{coifman2005geometric}.
%
When the normalized kernel is defined as
\begin{equation} \label{eq:norm_kernel}
\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1},
\end{equation}
the limiting diffusion operator is the Laplace-Beltrami operator \cite{nadler2006diffusion}, given by $\widetilde{\mathcal{L}} = \Delta$.
%
In other words, this diffusion captures only the geometry of the observations and is invariant to the potential (or, alternatively,
to the details of the sampling density).
This kernel is similar to the kernel in \eqref{eq:kernel}, but $D$ appears with coefficient -1 rather than $-1/2$.

Since local coordinates are based on the noise term in \eqref{eq:Langevin}, sets of measurements from different regions
will essentially detect the same system of coordinates (the same variable ``axes");
yet different sampling densities (arising, for example,
from the drift term in \eqref{eq:Langevin}) will result in different {\em scalings} of these axes.
%
Applying the Laplace-Beltrami normalization in the NIV context will thus eliminate sampling density effects.
%
An alternative approach would find the scaling transformations that optimally ``register"
data sampled from common regions (in the spirit of histogram reweighting \cite{ferrenberg1988new}).

@@

A na\"{\i}ve way to extend $\mathbf{f}$ to a new data point $\Psi(\tilde{i})$ is find the point's nearest neighbors in NIV space and average their function values.
%
A different, point-wise adaptive approach is described by Buchman {\em et al.} \cite{buchman2011high}:
high-dimensional hurricane tracks were estimated from low dimensional embedding coordinates using a weighted average of the points close to $\Psi(\tilde{i})$ in the embedded space.
%
However, this point-wise adaptation requires setting the nearest neighborhood radius parameter for every point.
%
The LP algorithm finds the appropriate nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
%
This radius will be large in smooth regions of the function, and small in regions in which $\mathbf{f}$ contains higher frequency components.
%
The LP approximation of a new, high dimensional point $\mathbf{Y}(\tilde{i})$ is calculated by a weighted average of the function values that belong to the neighboring points.
%
The weights are based on the pairwise distances in the intrinsic, low-dimensional space.
%
In practice, a set of smoothing operators $P_0, P_1, \ldots, P_l$, with
\begin{equation} \label{eq:LP_multi_scale_app}
P_l \propto e^{-d^2(\Psi(i),\Psi(j)) / 2^{-l} \sigma_0},
\end{equation}
are constructed and later extended to create the multiscale approximations as defined in \eqref{eq:LP_multi_scale}.
%
The LP algorithm for the inverse mapping is summarized in Algorithm \ref{algo_LP}.

@@

\item
Use the smoothing operators to obtain a multiscale representation (see \eqref{eq:LP_multi_scale_app}) of $\mathbf{f}:\Psi(i) \rightarrow Y(i)$.

@@

@@

We first consider a chemical reaction network involving  multiple enzyme-substrate interactions \cite{zagaris2012stability}.
%
The reaction steps that comprise the network are\\
\begin{equation}
\begin{array}{rcl}
E + S \overset{e_1}{\underset{e_{-1}}{\leftrightharpoons}} & E:S & \overset{e_2}{\rightarrow} E + S^{*} \\
S + E \overset{b_1}{\underset{b_{-1}}{\leftrightharpoons}} & S:E & \overset{b_2}{\rightarrow} S + E^{*}\\
D + S^{*} \overset{d_1}{\underset{d_{-1}}{\leftrightharpoons}} & D:S^{*} & \overset{d_2}{\rightarrow} D + S\\
F + E^{*} \overset{f_1}{\underset{f_{-1}}{\leftrightharpoons}} & F:E^{*} & \overset{f_2}{\rightarrow} F + E
\end{array}
\end{equation}
The ``$^{*}$'' denotes an activated form of a species, and the ``:'' denotes a complex formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.
%
There are 10 species in this reaction system.
%
However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions
(which we order as $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
%
We consider a parameter regime in which the ODE approximation of this scheme exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
%
Details about the specific parameter values can be found in Appendix \ref{app:rxn}.

@@

We generate 3000 initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$ uniformly at random from the
region of state space where all concentrations (respecting conservation laws) are non-negative.
%
We evolve each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain a point $\mathbf{Y}(t) \in \mathbb{R}^6$;
according to the time scales calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
but not long enough for the points to converge to a one-dimensional curve or to the final steady state (see Appendix \ref{app:rxn} for more details).
%
In our stochastic simulations, the initial points appear to converge to an approximate two-dimensional manifold
(in expected value, see Figure \ref{fig:rxn_manifolds}).
%
We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points ``on" this apparent two-dimensional manifold.
%
From each manifold point $\mathbf{y} \in \mathcal{Y}$, we run 20 short simulation ``bursts'', each for 0.2 time units.
%
We denote the endpoints from the short simulations as $\mathcal{Y}^{burst}(\mathbf{y})$.

@@

To determine whether a sufficient amount of data is available to obtain an accurate embedding, we compute the NIV embeddings for different numbers of data points
(arising from different numbers of trajectories initialized randomly in the full concentration space).
%
Figure \ref{fig:rxn_convergence} shows the two-dimensional NIV embeddings and corresponding eigenvalue spectra of the kernel for the Gillespie SSA example computed from 10, 100, 500, and 1500 data points.
%
We observe that the eigenvalue spectra appear converged for 100 data points and above.
%
This implies that a sufficient amount of data is available merely to discover the intrinsic variables.
%
However, we observe that 100 points are insufficient to construct a self-consistent embedding {\em in the two variables discovered}.
%
In order to account for the different scaling of the NIV coordinate axes, we require at least 500 data points.
%
This last convergence is exemplified in Figure \ref{fig:rxn_convergence} (c) and (d).
%
The convergence of the spectrum of the kernel and the ultimate self-consistency of the NIV embedding is our empirical indicator that a sufficient amount of data is available.
\begin{figure*}[ht]
    \includegraphics[width=6in]{fig4b}
    \caption{Two-dimensional NIV embeddings (top) and corresponding eigenvalue spectra (bottom) calculated from subsets of $\mathcal{Y}_1$ containing (a) 10 data points, (b) 100 data points, (c) 500 data points, and (d) 1500 data points. One can see that the embedding and spectrum in (a) is inconsistent with the other embeddings, and that, although the spectrum in (b) appears converged, the embedding is not converged until (c). }
    \label{fig:rxn_convergence}
\end{figure*}

@@

Our second example comes from the molecular dynamics simulation of a small peptide fragment.
%
Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein caricature for simulation studies
\cite{apostolakis1999calculation, bolhuis2000reaction, chekmarev2004long, ma2005automatic, frewen2009exploration, ferguson2011integrating}.
%
We simulate the motion of Ala2 in explicit solvent using the AMBER 10 molecular simulation package \cite{case2008Amber} with an
optimized version \cite{best2009optimized} of the AMBER ff03 force field \cite{duan2003point}.
%
The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}
with periodic boundary conditions, and the particle mesh Ewald method is used for long-range electrostatic interactions \cite{essmann1995smooth}.
%
The simulation is performed at constant volume and temperature (NVT ensemble), with the temperature being maintained at 300~K with a Langevin thermostat \cite{loncharich1992langevin}.
%
Hydrogen bond lengths are fixed using the SHAKE algorithm \cite{ryckaert1977numerical}.
%
The two dihedral angles $\phi$ and $\psi$ are known to parameterize the free energy surface, which contains three important minima (labeled A, B, and C, see Figure \ref{fig:ala_fes}).
%
Our simulations are concentrated around minimum B in the free energy surface, located at $\phi \approx -65^{\circ}$, $\psi \approx 150^{\circ}$.
%
We choose to sample only around minimum B, as each data set must sample the same region of the potential energy surface to obtain consistent NIV, as described in Section \ref{sec:diff_limit}.
%
We start many simulations at $10^{\circ}$ away from the minimum, and allow the simulations to each run for 0.1~ps, while recording the configuration of Ala2 every 1~fs (therefore, each trajectory is 100 points long).
%
Configurations are recorded with all atoms except the hydrogens.

@@

We first compare NIV with direct diffusion maps \cite{coifman2005geometric}, an established nonlinear dimensionality reduction technique.
%
We consider 10,000 data points from our simulation $\mathbf{Y}(1), \dots, \mathbf{Y}(10000) \in \mathbb{R}^{30}$; every 100 data points comes from a continuous simulation trajectory.
%
We construct two data sets:
%
$\mathcal{Y}_{even} = \{\mathbf{Y}(t) \text{ restricted to atoms 2, 4, 6, 8, and 10}: t=1, \dots, 10000 \}$,
and $\mathcal{Y}_{odd} = \{\mathbf{Y}(t) \text{ restricted to atoms 5, 7, and 9}: t=1, \dots, 10000 \}$ (see Figure \ref{fig:ala_fes} for the atom indexing).
%
We then compute the NIV and diffusion maps embeddings for $\mathcal{Y}_{even}$ and $\mathcal{Y}_{odd}$;
for NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$.

@@

@@

\begin{figure*}[ht]
    \centering
    \includegraphics[width=3in]{fig10b}
    \caption{The error in the reconstructed position of the z-coordinate of atom 10 as a function of $\psi_1$. One can see that the error is large when $\psi_1$ is large (similar to the error structure in Figure \ref{fig:LP_ex}), demonstrating that our data is indeed multiscale and that Laplacian Pyramids is an appropriate interpolation algorithm to use.}
    \label{fig:LapPyr_ala2_errors}
\end{figure*}

@@

%\bibliography{references}

@@

\begin{thebibliography}{46}%

@@

\bibitem [{\citenamefont {Bowen}, \citenamefont {Acrivos},\ and\ \citenamefont
  {Oppenheim}(1963)}]{bowen1963singular}%
  {Bowen}}, \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont {Acrivos}}, \
